---
title: "coding_exercise_TCGA_infiltration"
author: "Benedek Danko"
date: "2024-02-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/home/benedek_danko/R/coding_exercise_TCGA_infiltration")
```

```{r message=F, warning=F}
library("tidyverse")
library("ComplexHeatmap")
library("viridis")
```

# Coding exercise using TCGA immune infiltration data

Instructions are from [coding_exercise_TCGA_infiltration](https://github.com/immunitastx/coding_exercise_TCGA_infiltration).

First load the immune infiltration data:

```{r message=F, warning=F}
# CIBERSORT relative immune cell fractions:
tcga_immunfrac_cibersort <- read.table("data/TCGA.Kallisto.fullIDs.cibersort.relative.tsv",
                                       sep="\t", stringsAsFactors = F, header = T) %>% 
  mutate(SampleID=str_replace_all(SampleID, "\\.", "-"),
         SampleID=str_extract(SampleID, "TCGA-[\\d\\w]{2}-[\\d\\w]{4}-[\\w\\d]{2}"))
# Estimated absolute leukocyte fractions:
tcga_all_leuk <- read.table("data/TCGA_all_leuk_estimate.masked.20170107.tsv",
                                       sep="\t", stringsAsFactors = F, header = F) %>% 
  dplyr::rename("CancerType"="V1", "SampleID"="V2", "Fraction"="V3") %>% 
  mutate(SampleID=str_extract(SampleID, "TCGA-[\\w\\d]{2}-[\\w\\d]{4}-[\\w\\d]{2}"))
# Downloaded mutational data for LUAD:
tcga_luad_mut <- read.table("/home/benedek_danko/R/coding_exercise_TCGA_infiltration/data/TCGA_LUAD_SNV_harmonized.txt", 
                            header = T, sep="\t", stringsAsFactors = F, 
                            quote="", fill=FALSE) %>% 
  mutate(SampleID=str_extract(Tumor_Sample_Barcode, "TCGA-[\\w\\d]{2}-[\\w\\d]{4}-[\\w\\d]{2}"))
```

Check data:

```{r message=F, warning=F}
dim(tcga_immunfrac_cibersort)
dim(tcga_all_leuk)

str(tcga_immunfrac_cibersort)
str(tcga_all_leuk)
```

## Demonstrate that `TCGA.Kallisto.fullIDs.cibersort.relative.tsv` is relative immune cell abundance

```{r message=F, warning=F}
tcga_cibersort_rowsums <- rowSums(tcga_immunfrac_cibersort[, 3:24])
summary(tcga_cibersort_rowsums)
```

All cell tpye fractions sum up to 1 per sample, meaning that they are relative fractions (CIBERSORT).

## Calculate the absolute immune infiltration by using the `Leukocyte Fraction` using `TCGA_all_leuk_estimate.masked.20170107.tsv`

```{r message=F, warning=F}
# Calculate samples with more than 1 replicate:
tcga_replicate_samples <- tcga_immunfrac_cibersort %>% 
  group_by(SampleID) %>% 
  summarize(count=n()) %>% 
  filter(count > 1) %>% 
  arrange(desc(count)) 
nrow(tcga_replicate_samples)
```

There are 93 samples with replicates. Let's calculate mean fractions per sample for each cell type:

```{r message=F, warning=F}
tcga_immunfrac_cibersort_clean <- tcga_immunfrac_cibersort %>%
  group_by(SampleID) %>%
  summarise(across(
    starts_with("B.cells.naive"):ends_with("Neutrophils"),
    mean
  ))

tcga_all_leuk_clean <- tcga_all_leuk %>%
  dplyr::group_by(SampleID) %>%
  summarize(Fraction=mean(Fraction))
```

Calculate absolute fractions:

```{r message=F, warning=F}
# Intersect of the samples from the two tables:
intersect_samples <- intersect(tcga_immunfrac_cibersort_clean$SampleID,
                               tcga_all_leuk_clean$SampleID)
# Calculate absolute fractions:
tcga_abs_leuk_fractions <- tcga_immunfrac_cibersort_clean %>% 
  left_join(tcga_all_leuk_clean %>% 
              dplyr::select(c(SampleID, Fraction)), 
            by="SampleID") %>% 
  filter(SampleID %in% intersect_samples) %>% 
  filter(!is.na(Fraction)) %>% 
  mutate(Fraction=ifelse(Fraction < 0, 0, Fraction)) %>% 
  mutate(across(-c(SampleID, Fraction), ~ . * Fraction))
summary(rowSums(tcga_abs_leuk_fractions[,2:23]))
```

I set absolute fraction for values below 0 to 0. I also removed samples without absolute leukocyte measurements.

## Group similar immune cell types together and calculate the absolute immune infiltration for B_cells, T_cells, NK_cells, Monocytes, Macrophages, DCs, Mast, Eosinophils and Neutrophils

```{r message=F, warning=F}
tcga_abs_leuk_fractions_grouped <- tcga_abs_leuk_fractions %>% 
  dplyr::select(-Fraction) %>% 
  column_to_rownames("SampleID") %>% 
  t() %>% 
  as.data.frame() %>% 
  rownames_to_column("CellType") %>% 
  mutate(CellType=case_when(str_detect(CellType, "B.cells") ~ "B_cells",
                            str_detect(CellType, "T.cells") ~ "T_cells",
                            str_detect(CellType, "NK.cells") ~ "NK_cells",
                            str_detect(CellType, "Monocytes") ~ "Monocytes",
                            str_detect(CellType, "Macrophages") ~ "Macrophages",
                            str_detect(CellType, "Dendritic.cells") ~ "DCs",
                            str_detect(CellType, "Mast.cells") ~ "Mast_cells",
                            CellType %in% c("Eosinophils", "Neutrophils") ~ "Eosinophils_Neutrophils",
                            TRUE ~ CellType)) %>% 
  group_by(CellType) %>% 
  summarise(across(.cols = everything(), .fns = sum))
```

## Plot the immune infiltration of T cells across all cancer types in a boxplot and order the box by the median level of T cell infiltration from high to low

```{r message=F, warning=F, fig.width=12}
# Format data for T-cells only, add cancer type info:
tcga_t_cells_per_cancertype <- tcga_abs_leuk_fractions_grouped %>% 
  as.data.frame() %>% 
  column_to_rownames("CellType") %>% 
  t() %>% 
  as.data.frame() %>% 
  dplyr::select("T_cells") %>% 
  rownames_to_column("SampleID") %>% 
  left_join(tcga_immunfrac_cibersort %>% 
              dplyr::select(c(SampleID, CancerType))) %>% 
  group_by(CancerType) %>% 
  mutate(mean_T=mean(T_cells)) %>% 
  arrange(desc(mean_T)) 
tcga_t_cells_per_cancertype$CancerType <- factor(tcga_t_cells_per_cancertype$CancerType,
                                                 levels=unique(tcga_t_cells_per_cancertype$CancerType))

# Plot data:
ggplot(tcga_t_cells_per_cancertype,
       aes(x=CancerType, y=T_cells))+
  theme_bw()+
  geom_jitter(width = 0.25, alpha=.4)+
  geom_boxplot(outlier.shape = NA, width=.3, alpha=1)+
  xlab("Cancer type")+
  ylab("T-cells absolute fraction")
```

## Get the median level of immune infiltration of B_cells, T_cells, NK_cells, Monocytes, Macrophages, DCs, Mast and Eosinophils per cancer type and plot them in a heatmap

```{r message=F, warning=F, fig.height=8}
# Format data, calculate mean fractions per cancer type:
tcga_main_cells_per_cancertype <- tcga_abs_leuk_fractions_grouped %>% 
  as.data.frame() %>% 
  column_to_rownames("CellType") %>% 
  t() %>% 
  as.data.frame() %>% 
  rownames_to_column("SampleID") %>% 
  left_join(tcga_immunfrac_cibersort %>% 
              dplyr::select(c(SampleID, CancerType))) %>% 
  dplyr::select(-SampleID) %>% 
  group_by(CancerType) %>% 
  summarise(across(.cols = everything(), .fns = mean)) %>% 
  as.data.frame() %>% 
  column_to_rownames("CancerType") 

# Plot:
colnames(tcga_main_cells_per_cancertype) <- c("B-cells", "DCs", "Eosinophils/Neutrophils",
                                              "Macrophages", "Mast-cells", "Monocytes",
                                              "NK-cells", "Plasma-cells", "T-cells")
Heatmap(tcga_main_cells_per_cancertype,
                         cluster_rows = T,
                         cluster_columns = T,
                         col = rev(viridis(n=100, option="magma")),
                         show_column_names = T,
                         name="Mean absolute fraction",
                         border = TRUE,
                         show_row_names = T,
                         show_row_dend = T,
                         row_names_side = "left",
                         width=unit(100, "mm"))
```

## Plot the absolute infiltration level for B_cells, T_cells, NK_cells, Monocytes, Macrophages, DCs across different EGFR mutants for LUAD

Based on the LUAD masked mutational data downloaded using `TCGAbiolinks` and matched to our cell type abundance data, there are 28 Exon19 del., 17 Exon20 ins., 20 pL858R mut, 59 Other missense mut., and 472 WT cases. 

```{r message=F, warning=F, fig.width=12, fig.height=10}
# Get absolute fractions of main cell types for LUAD:
tcga_main_cells_luad <- tcga_abs_leuk_fractions_grouped %>% 
  as.data.frame() %>% 
  column_to_rownames("CellType") %>% 
  t() %>% 
  as.data.frame() %>% 
  rownames_to_column("SampleID") %>% 
  left_join(tcga_immunfrac_cibersort %>% 
              dplyr::select(c(SampleID, CancerType))) %>% 
  filter(CancerType=="LUAD") %>% 
  dplyr::select(-CancerType) 

# Get EGFR mutational data, create groups:
tcga_luad_egfr_mut <- tcga_luad_mut %>% 
  filter(Hugo_Symbol=="EGFR") %>% 
  mutate(group=case_when(Variant_Type=="INS" & Exon_Number=="20/28" ~ "exon20_ins",
                         Variant_Type=="DEL" & Exon_Number=="19/28" ~ "exon19_del",
                         HGVSp_Short == "p.L858R" ~"pL858R_mut",
                         TRUE ~ NA),
         group=ifelse(is.na(group) & Variant_Classification=="Missense_Mutation",
                      "other_missense_mut",
                      group)) %>% 
  dplyr::select(c(SampleID, group))

# Merge the two data frames (we assume that samples without EGFR mutational data are WT):
tcga_main_cells_luad_groups <- tcga_main_cells_luad %>% 
  left_join(tcga_luad_egfr_mut, by="SampleID") %>% 
  mutate(group=ifelse(is.na(group), "WT", group)) %>% 
  pivot_longer(-c(SampleID, group), names_to="CellType", values_to="Abs_abundance") %>% 
  mutate(CellType=factor(CellType, levels=c("T_cells", "B_cells", "Plasma.cells","NK_cells",
                                            "Macrophages", "Monocytes","DCs", "Eosinophils_Neutrophils",
                                            "Mast_cells"),
                         labels=c("T cells", "B cells", "Plasma cells", "NK cells",
                                  "Macrophages", "Monocytes", "DCs", "Eosinophils/Neutrophils",
                                  "Mast cells")),
         group=factor(group, levels=c("WT", "exon20_ins", "exon19_del",
                                      "pL858R_mut", "other_missense_mut"),
                      labels=c("WT", "Exon20 mut.", "Exon19 del.",
                               "pL858R mut.", "Other missense mut.")))

# Plot data:
ggplot(tcga_main_cells_luad_groups,
       aes(x=group, y=Abs_abundance))+
  theme_bw()+
  theme(axis.text.x = element_text(angle=60, hjust = 1))+
  geom_jitter(width = 0.25, alpha=.4)+
  geom_boxplot(outlier.shape = NA, width=.3, alpha=1)+
  facet_wrap(vars(CellType), scales="free")+
  xlab("EGFR mutational group")+
  ylab("Absolute abundance")
```